---
title: "R Data Prep"
author: "Ryan Geier"
date: "2023-04-10"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Set wd
```{r}
knitr::opts_knit$set(root.dir = normalizePath("C:/Users/USERNAME/FOLDER"))
```

#Load CSV to a dataframe, "read.table" {utils} R Documentation
```{r}
df <- read.csv("C:/Users/USERNAME/FOLDER/FILENAME.csv", stringsAsFactors = FALSE)
```

#Load Excel to a dataframe, with password included
```{r}
library(excel.link)
df <- xl.read.file("filename.xlsx", password = "password123", write.res.password = "password123")

#OR

library(readxl)
df <- read_excel("df.xlsx", sheet = "sheetname")

#NOTE: Have to set working directory prior
```



#Relocate column within the dataframe
```{r}
library(dplyr)
df <- df %>% relocate(columnname, .before = column2)
```

#Select columns 46:142 inclusive, drops all other columns. Use "-" before "c" to subtract columns 46:142. 
```{r}
df <- df[ , c(46:142)]

#OR 

library(dplyr)
df2 <- df %>% select(-(column1:column2))

#NOTE: Remove "-" to select columns "column1" to "column2"

#OR

df2 <- subset(df, select = -c(column1, column2, column3))

#OR 

num_cols <- unlist(lapply(df, is.numeric))
data_num <- df[ , num_cols]
```

#Remove columns by name
```{r}
df <- within(df, rm(column1, column2, column3))
```

#Complete Cases (remove NA's)
```{r}
df <-  df[complete.cases(df), ]

#OR

df <- na.omit(df)
```

#Complete Cases (remove NA's) by single column
```{r}
df <- df[!is.na(df$col_name), ]

#OR
subset(df, !is.na(col_name))

#OR

library(tidyr)
df %>% drop_na(col_name)
```

#Replace empty cells with NA
```{r}
df[df == ""] <- NA
```

#Impute NA's with FUN=mean (functions are FUN! :) )
```{r}
df2 <- df %>% 
  mutate_if(is.numeric, ~replace_na(.,mean(., na.rm = TRUE)))
```

#Select subset of rows by single columns value
```{r}
df <- df[df$columnname == "VALUE", ]

#OR

df2 <- df[!(df$columnname == "VALUE" | df$column2 == "VALUE2"), ]

#NOTE: Remove "!" to select those values, rather than exclude

#OR

df2 <- df[which(df$columnname == "VALUE"), ]

#OR

df2 <- subset(df, df$columnname == 1)
```

#Transform a single column's data type
```{r}
df$columnname <- as.factor(df$columnname)
```

#Transform column to date format
```{r}
df$DATE<-as.POSIXct(df$DATE, format="%m/%d/%Y")
```

#Transform multiple column's data type using lapply
```{r}
df[] <- lapply(df, as.numeric)

#OR

df[,c(46:135)] <- lapply(df[ , c(46:135)], as.numeric)
```

#Make NA's 0
```{r}
df[is.na(df)] <- 0
```

#Transform a binary factor column to a 0 or 1
```{r}
df$columnname <- 1 * (df$columnname == "some value")
```

#Drop unused factor levels
```{r}
df$columnname <- droplevels(df$columnname)
```

#Drop factor levels by name
```{r}
df <- df[!df$columnname %in% c("0", "1", "2", "3", "C7"), ]

#OR

df <- df[!df$columnname == "NULL", ]
```

#Recode/Combine factors
```{r}
df$columnname <- recode(df$columnname, 'c("0", "2", "3", "4", "5", "6", "NULL") = "0"; c("1") = "1";')
```

#Cut - Create factor from numerical data
```{r}
df$columncut <- cut(df$`columnname`, c(0.00,0.24,0.48,0.72,0.96,inf))

#OR

# create bins of size 1
df$column.bin <- .bincode(df$column, c(1:9)) 
```


#Create reference level for factors (useful for coefficient interpretation)
```{r}
df$columnname <- relevel(df$columnname, ref = "0")
```

#Combine columns using paste
```{r}
df$newcolumn <- paste(df$column1, df$column2, df$column3, df$column4, sep="_")

#Results in "'column1value'_'column2value'_'column3value'_'column4value'"
```

#Combine columns to create a new column
```{r}
df$newcolumn <- with(df, df$column1 + df$column2)

#OR

df$newcolumn <- df[, 89] + df[, 90] + df[, 91]
```


#Combine two data frames
```{r}
df <- cbind(df, df2)

#Or specify columns

df <- cbind(df, columnname = df2$columnname)

#NOTE: "columnname =" is a function to rename, else will become column "df2$columnname"
```

#Combine multiple data
```{r}
combine.df <- data.frame(df$column, df2$column2, df3, df4)
```


#Transform column(s) by function - Log+1 in example
```{r}
df[ , c(3:4)] <- log(df[ , c(3:4)] + 1)
```

#T-test
```{r}
t.test(x, y)

#OR on entire dataframe

library(reshape2)
meltdf <- melt(df)
pairwise.t.test(meltdf$value, meltdf$variable, p.adjust = "none")
```


#Correlation Matrix
```{r}
round(cor(df), 4)

#OR

library(tidyverse)
library(dplyr)

dfcorr <- df %>% 
  select_if(is.numeric) %>% 
  cor() %>% 
  round(digits = 2) %>% 
  as.data.frame()
dfcorr
```


#Heatmap using ggplot2
```{r}
library(ggplot2)
heatmap(cor(df), Rowv = NA, Colv = NA)

#OR

library(gplots)
heatmap.2(cor(df), Rowv = FALSE, Colv = FALSE, dendrogram = "none", 
          cellnote = round(cor(df),2), 
          notecol = "black", key = FALSE, trace = 'none', margins = c(2,2))

#OR

library(ggplot2)
library(reshape) # to generate input for the plot
cor.mat <- round(cor(df),2) # rounded correlation matrix
ret <- write.csv(x=cor.mat, file="df.csv")
melted.cor.mat <- melt(cor.mat)
ggplot(melted.cor.mat, aes(x = X1, y = X2, fill = value)) + 
  geom_tile() + 
  geom_text(aes(x = X1, y = X2, label = value))
```

#PLot using ggplot2
```{r}
library(ggplot2)
ggplot(data=df, aes(x=column1, y=column2)) + geom_point(size=2)+geom_smooth()

#OR

df$Color <-cut(df$columnname, breaks=c(-Inf, 49.9, Inf), labels=c("black", "green"))

ggplot(data=df) + geom_point(mapping=aes(x=column2, y=columnname), size=5, color=df$Color)

#OR

ggplot(df) + geom_point(aes(x = column1, y = column2)) + 
  scale_x_log10(breaks = 10^(-2:2),
                labels = format(10^(-2:2), scientific = FALSE, drop0trailing = TRUE)) +
  scale_y_log10(breaks = c(5, 10, 20, 40))
```

#Plot base R
```{r}
plot(x, y)

#OR

hist(x)

#OR

barplot(x, y)

#OR

boxplot(df$column1 ~ df$column2, xlab = "Column1", ylab = "Column2")

#OR

par(mar=c(5.1, 5.1, 4.1, 2.1))
par(xpd=TRUE)
plot(df$column1 ~ df$column2, xlab = "Column 1", ylab = "Column 2")
legend("topleft", inset=c(0, -0.2),
legend = c("Column 1", "Column 2 (in thousands of USD)"),
pch = 1, cex = 0.5)

#OR

data.for.plot <- aggregate(df$Retail.Price, by = list(df$Store.Postcode), FUN = mean)
names(data.for.plot) <- c("Column 1", "Column 2")
barplot(data.for.plot$column2, names.arg = data.for.plot$column1,
xlab = "Column 1", ylab = "Column 2")

#OR

# to use logarithmic scale set argument log = to either 'x', 'y', or 'xy'. 
plot(df$columnname ~ df$column2, 
     xlab = "column2", ylab = "columnname", log = 'xy')

#OR

boxplot(df$columnname ~ df$column2, 
        xlab = "column2", ylab = "columnname", log = 'y')
```

#Plot with Map - Google API
```{r}
library(ggmap)
seattle <- read.csv("C:/Users/USERNAME/FOLDER/filename.csv")
head(seattle)
#you will need to create a google account to get a GoodgleAPI key
register_google(key = "LongAlphaNumericKeyHere12345")

Map <- get_map("Seattle, WA", zoom = 11)
ggmap(Map) + geom_point(aes(x = Longitude, y = Latitude), data = seattle,
                        alpha = 0.4, colour = "red", size = 0.5)

```



#Aggregate by function (change FUN = mean to other function if necessary)
```{r}
aggregation <-aggregate(aggregatedvalue ~ aggregatedby, df, mean)

#View Aggregation

aggregation[with(aggregation, order(aggregatedvalue)),]

#OR

aggregation <- aggregate(df, by=list(cluster=df$columnname), mean)
```

#Table 
```{r}
table(var1, var2)

#OR

prop.tbl <- prop.table(table, margin=2)

#OR

#PIVOT TABLE EXAMPLE

library(reshape) 
# create bins of size 1
df$column.bin <- .bincode(df$column, c(1:9)) 

# use melt() to stack a set of columns into a single column of data.
# stack MEDV values for each combination of (binned) RM and CHAS
mlt <- melt(df, id=c("column.bin", "column2"), measure=c("column3"))

# use cast() to reshape data and generate pivot table
cast(mlt, column.bin ~ column2, subset=variable=="column3", 
     margins=c("grand_row", "grand_col"), mean)
```

#Univariate and Bivariate Loop print
```{r plots, echo=FALSE}
for (i in colnames(df %>% select (c(105:138)))) {
  a <- df[, i]
  b <- (class(a))
  c <- hist(a, xlab = i)
  cat(" \n")
  print(i)
  print("Univariate Summary")
  print(paste("data type:", b))
  print(summary(as.numeric(a)))
  print(c)
  cat(" \n")
  cat(" \n")
  cat(" \n")
  print("Bivariate Analysis")
  print(ggplot(df, aes_string(x = i, y = df$columnname)) + geom_point() + geom_smooth() + labs(y= "label_column_name"))  
  print(ols_regress(paste("columnname ~", i), data = df))
  print(summary(lm(paste("columnname ~", i), data = df)))
  cat(" \n")
  cat(" \n")
  cat(" \n")
  }
```

#Create Matrix, Sparse Matrix, from dataframe
```{r}
mx <- data.matrix(df)
mx <- Matrix(mx, sparse=TRUE)
```

#Sparse Matrix - Partitioned
```{r}
train.sparse <- sparse.model.matrix(~., train.df[1:90])
test.sparse <- sparse.model.matrix(~., valid.df[1:90])
```

#Partition Test and Train Data
```{r}
set.seed(2)
train.index <- sample(c(1:dim(df)[1]), dim(df)[1]*0.7)
train.df <- df[train.index, ]
valid.df <- df[-train.index, ]
```




